Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-06-22 src/tex/ms.tex

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), $f_{\mathrm{max}}$ and $f_{\mathrm{min}}$ are
the relevant maximum and minimum frequencies for the integration.
We note here that the mismatch can be viewed as the mean square error (MSE) between the
two waveforms.

\te{Up to here}

Since we wish to optimize the model over the whole parameter space, we need to compare multiple model generated waveforms with NR waveforms. However, mismatch only quantifies the difference between IMRPhenomD and NR waveform for one particular set of intrinsic parameters. To take into account of various different waveforms in the parameter space, we pick waveforms from the different parts of the parameter space. We define the loss function as an average of training waveforms in two ways, the simple average of mismatches and the normalize average of mismatches,  
\begin{align}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{nl}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of individual training waveforms.
Note that we choose to use two different averages, since they have different 
preferences in optimization base on waveform mismatches. For the first choice, 
simple average serves as the simplest choice of loss function, but is prone to 
be dominated by a single waveform with a large mismatch. Other waveforms with 
smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such optimization. Alternatively, the second choice, normalized 
average eliminates the aforementioned issue. Nevertheless, it excludes the
information on initial mismatches. $\mathcal{L}_{\mathrm{nl}}$ restricts every
training waveform to decrease at similar rates, hence it is hard to obtain
optimized waveforms with mismatches in the same order of magnitude. Instead,
their ratios in mismatches would remain approximately the same. Conversely,
$\mathcal{L}_{\mathrm{ave}}$ allows the loss function to automatically adjust
and individual mismatches would be in a similar order of magnitude after 
optimization. In this paper, we showcase the results of using both loss functions 
and examine the differences between them.

@@

We point out that our treatment to NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}. In the original calibration process, training waveforms are hybrid waveforms of NR and SpinAlignedEOB
(SEOB) waveforms. The low frequency inspiral part is taken from the SEOB
waveforms while the rest of the waveforms are taken from NR simulations.
Instead, we solely use NR waveforms for comparison, since we are only exploring the possibility of optimizing waveform models. Thus, for simplicity sake, we ignore this procedure. On the other hand, most NR waveforms used (for both training and testing) have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, in
which they are long enough to contain part of the inspiral segment and all
merger and ringdown frequency information. We take the frequency limits as
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set higher
than that in the original calibration due to NR length. When compared with
IMRPhenomC, the frequency range is slightly extended to have a higher maximum
frequency \citep{santamaria2010matching}. We have the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$, we see the validation loss in 
Fig.~\ref{fig:loss} plateau at around $N=12000$, hence we end the optimization here.  
%\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

>>>>

In order to optimize the coefficients, we need to define a loss function that
quantifies the difference between the waveform model and the target NR simulations which we want to match.
Here we adopt a quantity commonly used in GW physics called the \textit{mismatch} function \citep{Owen:1995tm,husa2016frequency}. 
It is defined as
\begin{align} \label{eq:mismatch}
	\mathcal{M}(h_1, h_2)=1-\max_{t_0, \phi_0}\langle \hat{h}_1, \hat{h}_2\rangle,
\end{align}
where $h_{1,2}$ are the two GW waveforms we are comparing, and $t_0$ and $\phi_0$
are a relative time and phase shift respectively. 
The inner product, $\langle h_1, h_2 \rangle$, is defined as 
\begin{align}\label{eq:inner_prod}
	\langle h_1, h_2 \rangle = 4\Re\int_{f_{\mathrm{min}}}^{f_{\mathrm{max}}}\frac{h_1(f)h_2^{\ast}(f)}{S_n(f)}\,df,
\end{align}
where $\hat{h}=h/\sqrt{\langle h, h \rangle}$ is the normalized GW strain,
$S_n(f)$ is the power spectral density (PSD), and $f_{\mathrm{max}}$ ($f_{\mathrm{min}}$) is
the maximum (minimum) frequency for the integration.
We note here that the mismatch can be viewed as the mean square error (MSE) between the
two waveforms.

Since we wish to optimize the model over the whole parameter space,
we need to compare multiple model generated waveforms with NR waveforms.
However, the mismatch is only defined for two input waveforms at a particular set of intrinsic parameters.
We therefore define the loss function as an average of training waveforms in two ways,
the simple average of mismatches and the normalized average of mismatches,  
\begin{align}
	\mathcal{L}_{\mathrm{ave}}&=\frac{1}{N}\sum_{i=1}^N\mathcal{M}_i \\
	\mathcal{L}_{\mathrm{nl}}&=\frac{1}{N}\sum_{i=1}^N\frac{\mathcal{M}_i}{\mathcal{M}_{i,\mathrm{ini}}},
\end{align}	
where $\mathcal{M}_i$ represents the mismatch of an individual training waveform,
$\mathcal{M}_{i,\mathrm{ini}}$ represents the initial mismatch of the individual
training waveform, and $N$ is the total number of training waveforms.

The two loss functions are chosen to give different 
behaviour during the optimization process.
The simple average, $\mathcal{L}_{\mathrm{mean}}$, serves as the simplest choice of loss function,
but is prone to be dominated by a single point in parameter space with a large mismatch.
Other points with smaller mismatches would be insignificant comparatively, and might not be able 
to improve under such a loss function.
Alternatively, the normalized average, $\mathcal{L}_{\mathrm{norm}}$, eliminates the aforementioned issue
by encouraging the waveform to improve at each training point at a similar rate.
The ratio in $\mathcal{L}_{\mathrm{norm}}$ will therefore remain approximately the same for each training point.
Conversely, $\mathcal{L}_{\mathrm{mean}}$ allows the loss function to automatically adjust
and preferencially optimize the largest mismatches, encouraging the waveform to have similar mismatches everywhere.
In this paper, we show the results of using both loss functions 
and examine the differences between them.

@@

We point out that our treatment of NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}.
In the original calibration process, the training waveforms are hybrids of NR and SpinAlignedEOB (SEOB) waveforms.
The low frequency inspiral part is taken from the SEOB
waveforms while the rest is taken from NR simulations.
Instead, we solely use NR waveforms for calibration since most NR waveforms used (for both training and testing)
have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, to contain part of the inspiral segment and all
merger and ringdown frequency information. We use the frequency limits
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set slightly higher
than in the original calibration due to the NR simulation length.
When compared with IMRPhenomC, the frequency range is slightly extended to have a higher maximum
frequency \citep{santamaria2010matching}.
We use the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of the GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency} because they likely lie in the
neighborhood of the minimum that we wish to find. 
We fix our learning rate, $\alpha$, to be $=10^{-6}$, which is small enough to ensure we 
don't move far from the minimum.
Finally, we stop the optimization when the validation loss stops decreasing.
This can be seen in Fig.~\ref{fig:loss} at around $12000$ iterations.
